The Effect of Elections and Prime Ministers on Discussion in the Australian Federal Parliament (1901--2018)

### 2018-12-09 outputs/docs/events/paper.Rmd
Update intro and data

---
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    number_sections: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "The Effect of Elections and Prime Ministers on Discussion in the Australian Federal Parliament (1901--2018)"
thanks: "Thank you to Chris Cochrane, Dan Simpson, Jill Sheppard, John McAndrews, John Tang, Leslie Root, Martine Mariotti, Matt Jacob, Matthew Kerby, Myles Clark, Ruth Howlett, Tianyi Wang, Tim Hatton, and Zach Ward for their helpful suggestions; and to the UC Berkeley Demography Department for the use of their computing resources. We are grateful for the many excellent comments that we received from seminar participants at the ANU SPIR, the ANU RSE, the Australian Parliamentary Library, and the UofT Political Behavior Group. Comments and suggestions on the `r format(Sys.time(), '%d %B %Y')` version of this paper welcome at: rohan.alexander@anu.edu.au."
author:
- name: Monica Alexander
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: Australian National University
abstract: "A large part of politics consists of reacting to events. But not all events provoke a reaction and so understanding political behaviour requires examining which ones do and how this changes over time. We systematically analyse how discussion in the Australian Federal Parliament changes in response to two types of events: elections and changed prime ministers. To do this we first create a new dataset of what was said in the Australian Federal Parliament from 1901 through to 2018 based on available public records. We reduce the dimensionality of this dataset by using a correlated topic model, and then analyse the effect of these two types of events using a Bayesian hierarchical Dirichlet model that has several advantages over existing models. We find that: changes in prime minister tend to be associated with topic changes even when the party in power does not change; elections that do not result in a change in prime minister are rarely associated with topic changes; and that these effects have become more pronounced since the 1980s."
keywords: "text-as-data, Australian politics, unsupervised machine learning, Bayesian hierarchical Dirichlet model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
toc: FALSE
spacing: double
bibliography: ../bibliography.bib
biblio-style: apsr
endnote: no

@@

**Update first para** Maybe include 
"So maybe the implication here is something about the role of govt agenda setting in the chamber?"

In this paper we examine text records of what was said in the Australian Federal Parliament. We create a dataset based on the sitting days from 1901 through to 2018. We have records for 7,934 days in the House of Representatives and 6,746 days in the Senate. Our dataset consists of counts of every word that occurs at least thrice, after we convert all words to lower-case; remove names, numbers, and punctuation; join commonly co-located words; and remove stop words.

We systematically analyse the text in two stages, each of which takes advantage of a different statistical technique. First, to reduce the dimensionality of our dataset we use a topic model that allows correlation between topics. Second, we analyse those topics using a Bayesian hierarchical Dirichlet model to examine changes at various types of events. We explicitly consider two types of events: changes in prime minister and elections. After taking into consideration these effects, we can then explore other significant events.

We find that: firstly, changes in prime minister tend be associated with topic changes even when the party in power does not change. For instance, the change from Hughes to Bruce in 1922, Menzies to Holt in 1966; Hawke to Keating in 1991; and Rudd to Gillard in 2010 are all associated with significant changes in topics despite no change in the party in power. Secondly, as expected, elections where the party in power also changes, such as Fisher in 1910 and 1914, Menzies in 1949, Whitlam in 1972, Fraser in 1975, Hawke in 1983, Howard in 1996, Rudd in 2007, and Abbott in 2013 are associated with topic changes, but the 1974 Whitlam, 1980 Fraser, and the 1998 and 2004 Howard re-elections, stand out as elections where the prime minister did not change but there was a significant change in topics. Finally, economic events, such as financial crises, have less significant effects than other events such as terrorism.

As our dataset covers 118 years we are able to see how the effect of events changes over time. We find that the events that we are interested in are rarely significant in the first half of our dataset. With a small number of exceptions, even changes of prime minister where the party in power also changed were not associated with overly large changes in the topics of parliamentary discussion. If prime ministers try to more thoroughly distinguish themselves from their predecessor, then longer-term policy may be more difficult to enact. Similarly, there is the danger that larger-scale signature-accomplishments of a prime minister may be neglected by their successor for solely political reasons.

Our work contributes to the growing literature in quantitative social science that considers text as an input to statistical methods, rather than requiring separate analysis. This literature sits across and draws from various historically-separate disciplines including applied statistics, economics, and political science. We contribute to this literature in terms of both data and methods, as well as our key findings. From a data perspective we bring to bear an essentially-complete record of what was said in the Australian Federal Parliament on a daily basis, and we make our dataset available to other researchers via the R package \texttt{hansardr}. From a methods perspective, our analysis model has several advantages over existing methods. These include: allowing the events to have more-complicated auto-correlated functional forms; implementing pooling across groups of similar documents; and identifying outlying topic distributions without the need to pre-specify the event of interest. Those contributions allow us to to find that the effects in Australia of elections and prime ministers appears to have become more pronounced since the 1980s. As the digitisation of historical sources continues and computational power becomes cheaper, we expect the popularlity of quantitative approaches to text to further increase and the data and methods in our paper can be improved on in a variety of ways.

@@

Following the example of the UK a daily text record called Hansard of what was said in the Australian Federal Parliament has been made available since Federation in 1901.[^Verbatimfootnote] Analysing Hansard records and their equivalents directly is increasingly viable as new methods and reduced computational costs make it easier.

@@

The percentage of stop-words in each record is reasonably consistent over time. This suggests that there is no significant difference in the quality of the parsing over time, although manual inspection does suggest there is some improvement in quality over time. Details of this check are provided in Appendix \@ref(stopwordsgraph). We use Hansard records on a daily basis in this paper. We pre-process our text before applying a topic model. The specific steps that we take are to: remove numbers and punctuation; change the words to lower case; and concatenate multi-word names titles and phrases, such as new zealand to new_zealand. Then the sentences are de-constructed and each word considered individually. The resulting dataset used for analysis contains counts of words by day for each of the 14,680 sitting days between 1901 and 2018. We do not stem the words because @SchofieldMimno2016 suggest that there is no significant benefit.

**[Add discussion of how wonderful we are for making the data available.]**

>>>>

---
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    number_sections: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "The Effect of Elections and Prime Ministers on Discussion in the Australian Federal Parliament (1901--2018)"
thanks: "Thank you to Chris Cochrane, Dan Simpson, Jill Sheppard, John McAndrews, John Tang, Leslie Root, Martine Mariotti, Matt Jacob, Matthew Kerby, Myles Clark, Ruth Howlett, Tianyi Wang, Tim Hatton, and Zach Ward for their helpful suggestions; and to the UC Berkeley Demography Department for the use of their computing resources. We are grateful for the many excellent comments that we received from seminar participants at the ANU SPIR, the ANU RSE, the Australian Parliamentary Library, and the UofT Political Behavior Group. Comments and suggestions on the `r format(Sys.time(), '%d %B %Y')` version of this paper welcome at: rohan.alexander@anu.edu.au."
author:
- name: Monica Alexander
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: Australian National University
abstract: "A large part of politics consists of reacting to events, but not all events provoke a reaction. As such, understanding political behaviour requires examining the reaction to events and how this changes over time. We systematically analyse how discussion in the Australian Federal Parliament changes in response to two types of events: elections and changed prime ministers. To do this we first create a new dataset of what was said in the Australian Federal Parliament from 1901 through to 2018 based on available public records. We reduce the dimensionality of this dataset by using a correlated topic model, and then analyse the effect of these two types of events using a Bayesian hierarchical Dirichlet model that has several advantages over existing models. We find that: changes in prime minister tend to be associated with topic changes even when the party in power does not change; elections that do not result in a change in prime minister are rarely associated with topic changes; and that these effects have become more pronounced since the 1980s."
keywords: "text-as-data, Australian politics, unsupervised machine learning, Bayesian hierarchical Dirichlet model"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
toc: FALSE
spacing: double
bibliography: ../bibliography.bib
biblio-style: apsr
endnote: no

@@

Events are a well-known cause of political behaviour. But the extent of their effect is varied and changes over time. Here we focus on just two types of events: elections, and a change in prime minister. We examine the effect of these types of events on the topics that are discussed in the Australian Federal Parliament.

The main data source for our paper is the text record of what was said in the Australian Federal Parliament known as 'Hansard'. From this we create a dataset based on the sitting days from 1901 through to 2018. We have records for 7,934 days in the House of Representatives and 6,746 days in the Senate. Our dataset consists of counts of every word that occurs at least thrice, after we convert all words to lower-case; remove names, numbers, and punctuation; join commonly co-located words; and remove stop words.

We systematically analyse the text in two stages, each of which takes advantage of a different statistical technique. First, to reduce the dimensionality of our dataset we use a topic model that allows correlation between topics. Second, we analyse those topics using a Bayesian hierarchical Dirichlet model to examine changes at various types of events. We explicitly consider two types of events: changes in prime minister and elections. After taking these into consideration these effects, we can then explore other significant events.

We find that: firstly, changes in prime minister tend be associated with topic changes even when the party in power does not change. For instance, the change from Hughes to Bruce in 1922, Menzies to Holt in 1966; Hawke to Keating in 1991; and Rudd to Gillard in 2010 are all associated with significant changes in topics despite no change in the party in power. Secondly, as expected, elections where the party in power also changes, such as Fisher in 1910 and 1914, Menzies in 1949, Whitlam in 1972, Fraser in 1975, Hawke in 1983, Howard in 1996, Rudd in 2007, and Abbott in 2013 are associated with topic changes, but the 1974 Whitlam, 1980 Fraser, and the 1998 and 2004 Howard re-elections, stand out as elections where the prime minister did not change but there was a significant change in topics. The effects in Australia of elections and prime ministers appear to have become more pronounced since the 1980s. Finally, economic events, such as financial crises, have less significant effects than other events such as terrorism.

As our dataset covers 118 years we are able to see how the effect of events changes over time. We find that the events that we are interested in are rarely significant in the first half of our dataset. With a small number of exceptions, even changes of prime minister where the party in power also changed were not associated with overly large changes in the topics of parliamentary discussion. It may be that these days prime ministers are trying to more thoroughly distinguish themselves from their predecessor, or that the role of the government in agenda setting in the Australian Federal Parliament is changing.

Our work contributes to the growing literature in quantitative social science that considers text as an input to statistical methods, rather than requiring separate analysis. This literature sits across and draws from various historically-separate disciplines including applied statistics, economics, and political science. We contribute to this literature in terms of both data and methods, as well as our key findings. From a data perspective we bring to bear an essentially-complete record of what was said in the Australian Federal Parliament on a daily basis, and we make our dataset available to other researchers via the R package \texttt{hansardr}. From a methods perspective, our analysis model has several advantages over existing methods. These include: allowing the events to have more-complicated auto-correlated functional forms; implementing pooling across groups of similar documents; and identifying outlying topic distributions without the need to pre-specify the event of interest. As the digitisation of historical sources continues and computational power becomes cheaper, we expect the popularity of quantitative approaches to text to further increase and the data and methods in our paper can be improved on in a variety of ways.

@@

Following the example of the UK a daily text record called Hansard of what was said in the Australian Federal Parliament has been made available since Federation in 1901.[^Verbatimfootnote] Analysing Hansard records and their equivalents is increasingly viable as new methods and reduced computational costs make it easier.

@@

The percentage of stop-words in each record is reasonably consistent over time. This suggests that there is no significant difference in the quality of the parsing over time, although manual inspection does suggest there is some improvement in quality over time. Details of this check are provided in Appendix \@ref(stopwordsgraph). We use Hansard records on a daily basis in this paper. We pre-process our text before applying a topic model. The specific steps that we take are to: remove numbers and punctuation; change the words to lower case; and concatenate multi-word names titles and phrases, such as new zealand to new_zealand. We do not stem the words because @SchofieldMimno2016 suggest that there is no significant benefit. Then the sentences are de-constructed and each word considered individually. The resulting dataset used for analysis contains counts of words by day for each of the 14,680 sitting days between 1901 and 2018.[^datasourcefootnote]

[^datasourcefootnote]: We have made our dataset available via the R package \texttt{hansardr}. R users can draw on the devtools package [@Wickham2018] to install it from: https://github.com/RohanAlexander/hansardr. Users of other languages can download the daily CSV files from the data-raw folder of that repository.
